\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-05-26 scratch/notes_john/JSS-article/JSS-article-3.Rmd
further edits to draft JSS paper

[^recursive-cv]: What we call here "recursive CV" has also been termed "nested CV" [e.g., in the \pkg{cvms} package by @OlsenZachariae:2024]. We prefer, however, to reserve the term "nested CV" for the procedure described by [@BatesHastieTibshirani:2023].

In Section \ref{comparing-competing-models}, following @JamesEtAl:2021[Secs. 5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of `mpg` to `horsepower` for the `Auto` data, saving the 10 models, named `m.1` through `m.10`, in `mlist`. We then used `cv()` to compare the cross-validated MSE for the 10 models, discovering that the 7th degree polynomial had the smallest MSE (by a small margin).

If we then select the 7th degree polynomial model, intending to use it for prediction, the CV estimate of the MSE for this model will be optimistic. One solution is to cross-validate the process of using CV to select the "best" model---that is, to apply CV to CV recursively. The function `selectModelList()`, which is suitable for use with `cv()`, implements this idea.

@@

Some \proglang{R} packages for statistical learning include facilities for cross-validation. Notable examples are the \pkg{caret} package [@Kuhn:2008] and the \pkg{tidyfit} package [@Pfitzinger:2024]; the latter employs \pkg{rsample} [@FrickEtAl:2024] for CV, which we will discuss presently This contrasts with the approach taken in the \pkg{cv} package, which, as we have explained, directly supports classes of standard \proglang{R} statistical models.

@@

library(purrr, quietly=TRUE)
# apply CV:
rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
# summarize results:
rs_obj$accuracy <- map_dbl(rs_obj$results,
                           function(x) mean(x$correct))
# accuracy is the complement of the Bayes-rule CV criterion:
1 - summary(rs_obj$accuracy) 
```
[^k-fold-used-in-rsample]: The example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise the code from the vignette is only trivially modified here.

>>>>

[^recursive-cv]: What we call here "recursive CV" has also been termed "nested CV" [e.g., in the \pkg{cvms} package by @OlsenZachariae:2024]. We prefer, however, to reserve the term "nested CV" for the procedure described by @BatesHastieTibshirani:2023.

In Section \ref{comparing-competing-models}, following @JamesEtAl:2021[Secs. 5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of `mpg` to `horsepower` for the `Auto` data, saving the 10 models, named `m.1` through `m.10`, in the list `mlist`. We then used `cv()` to compare the cross-validated MSE for the 10 models, discovering that the 7th degree polynomial had the smallest MSE (by a small margin).

If we select the 7th degree polynomial model, intending to use it for prediction, the CV estimate of the MSE for this model will be optimistic. One solution is to cross-validate the process of using CV to select the "best" model---that is, to apply CV to CV recursively. The function `selectModelList()`, which is suitable for use with `cv()`, implements this idea.

@@

Some \proglang{R} packages for statistical learning include facilities for cross-validation. Notable examples are the \pkg{caret} package [@Kuhn:2008] and the \pkg{tidyfit} package [@Pfitzinger:2024]; the latter employs \pkg{rsample} [@FrickEtAl:2024] for CV, which we will discuss presently. This contrasts with the approach taken in the \pkg{cv} package, which, as we have explained, directly supports classes of standard \proglang{R} statistical models.

@@

library(purrr, warn.conflicts=FALSE)
# apply CV:
rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
# summarize results:
rs_obj$accuracy <- map_dbl(rs_obj$results,
                           function(x) mean(x$correct))
# accuracy is the complement of the Bayes-rule CV criterion:
1 - summary(rs_obj$accuracy) 
```
[^k-fold-used-in-rsample]: The example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise the code from the vignette is only trivially modified here.

### 2024-05-26 scratch/notes_john/JSS-article/JSS-article-3.tex
further edits to draft JSS paper

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 866542
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.09701
bias-adjusted cross-validation criterion = 19.09108
95% CI for bias-adjusted CV criterion = (15.65339, 22.52877)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\subsection[Example: Applying recursive CV to polynomial regression for
the Auto data]{\texorpdfstring{Example: Applying recursive CV to
polynomial regression for the Auto
data\footnote{What we call here ``recursive CV'' has also been termed
  ``nested CV'' \citep[e.g., in the \pkg{cvms} package
  by][]{OlsenZachariae:2024}. We prefer, however, to reserve the term
  ``nested CV'' for the procedure described by
  \citep{BatesHastieTibshirani:2023}.}}{Example: Applying recursive CV to polynomial regression for the Auto data}}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-datarecursive-cv}

In Section \ref{comparing-competing-models}, following \citet[Secs. 5.1,
5.3]{JamesEtAl:2021}, we fit polynomial regressions up to degree 10 to
the relationship of \texttt{mpg} to \texttt{horsepower} for the
\texttt{Auto} data, saving the 10 models, named \texttt{m.1} through
\texttt{m.10}, in \texttt{mlist}. We then used \texttt{cv()} to compare
the cross-validated MSE for the 10 models, discovering that the 7th
degree polynomial had the smallest MSE (by a small margin).

If we then select the 7th degree polynomial model, intending to use it
for prediction, the CV estimate of the MSE for this model will be
optimistic. One solution is to cross-validate the process of using CV to
select the ``best'' model---that is, to apply CV to CV recursively. The
function \texttt{selectModelList()}, which is suitable for use with
\texttt{cv()}, implements this idea.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function(y, yhat) {
  if (!all(y %in% c(0, 1)))
    stop("response values not all 0 or 1")
  if (any(yhat < 0) ||
      any(yhat > 1))
    stop("fitted values outside of interval [0, 1]")
  yhat <- round(yhat)
  result <- mean(y != yhat) # proportion in error
  attr(result, "casewise loss") <- "y != round(yhat)"
  result
}
<bytecode: 0x13b740948>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

Some \proglang{R} packages for statistical learning include facilities
for cross-validation. Notable examples are the \pkg{caret} package
\citep{Kuhn:2008} and the \pkg{tidyfit} package \citep{Pfitzinger:2024};
the latter employs \pkg{rsample} \citep{FrickEtAl:2024} for CV, which we
will discuss presently This contrasts with the approach taken in the
\pkg{cv} package, which, as we have explained, directly supports classes
of standard \proglang{R} statistical models.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library(rsample)
R> 
R> data("attrition", package = "modeldata")
R> nrow(attrition)
\end{CodeInput}
\begin{CodeOutput}
[1] 1470
\end{CodeOutput}
\begin{CodeInput}
R> # formula for model to be fit; Attrition is binary:
R> mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome
R> 
R> rs_obj <- loo_cv(attrition) # original uses vfold_cv()
R> 
R> # user-supplied function:
R> holdout_results <- function(splits, ...) {
+   mod <- glm(..., data=analysis(splits), family=binomial)
+   holdout <- assessment(splits)
+   res <- broom::augment(mod, newdata=holdout)
+   lvls <- levels(holdout$Attrition)
+   predictions <- factor(ifelse(res$.fitted > 0, lvls[2], lvls[1]),
+                         levels=lvls)
+   res$correct <- predictions == holdout$Attrition
+   res
+ }
R> 
R> library(purrr, quietly=TRUE)
\end{CodeInput}
\begin{CodeOutput}

Attaching package: 'purrr'
\end{CodeOutput}
\begin{CodeOutput}
The following object is masked from 'package:car':

    some
\end{CodeOutput}
\begin{CodeOutput}
The following objects are masked from 'package:foreach':

    accumulate, when
\end{CodeOutput}
\begin{CodeInput}
R> # apply CV:
R> rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
R> # summarize results:
R> rs_obj$accuracy <- map_dbl(rs_obj$results,
+                            function(x) mean(x$correct))
R> # accuracy is the complement of the Bayes-rule CV criterion:
R> 1 - summary(rs_obj$accuracy) 
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 1.0000  0.0000  0.0000  0.1612  0.0000  0.0000 
\end{CodeOutput}
\end{CodeChunk}

>>>>

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 968919
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.22048
bias-adjusted cross-validation criterion = 19.20805
95% CI for bias-adjusted CV criterion = (15.74739, 22.66871)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\subsection[Example: Applying recursive CV to polynomial regression for
the Auto data]{\texorpdfstring{Example: Applying recursive CV to
polynomial regression for the Auto
data\footnote{What we call here ``recursive CV'' has also been termed
  ``nested CV'' \citep[e.g., in the \pkg{cvms} package
  by][]{OlsenZachariae:2024}. We prefer, however, to reserve the term
  ``nested CV'' for the procedure described by
  \citet{BatesHastieTibshirani:2023}.}}{Example: Applying recursive CV to polynomial regression for the Auto data}}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-datarecursive-cv}

In Section \ref{comparing-competing-models}, following \citet[Secs. 5.1,
5.3]{JamesEtAl:2021}, we fit polynomial regressions up to degree 10 to
the relationship of \texttt{mpg} to \texttt{horsepower} for the
\texttt{Auto} data, saving the 10 models, named \texttt{m.1} through
\texttt{m.10}, in the list \texttt{mlist}. We then used \texttt{cv()} to
compare the cross-validated MSE for the 10 models, discovering that the
7th degree polynomial had the smallest MSE (by a small margin).

If we select the 7th degree polynomial model, intending to use it for
prediction, the CV estimate of the MSE for this model will be
optimistic. One solution is to cross-validate the process of using CV to
select the ``best'' model---that is, to apply CV to CV recursively. The
function \texttt{selectModelList()}, which is suitable for use with
\texttt{cv()}, implements this idea.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function(y, yhat) {
  if (!all(y %in% c(0, 1)))
    stop("response values not all 0 or 1")
  if (any(yhat < 0) ||
      any(yhat > 1))
    stop("fitted values outside of interval [0, 1]")
  yhat <- round(yhat)
  result <- mean(y != yhat) # proportion in error
  attr(result, "casewise loss") <- "y != round(yhat)"
  result
}
<bytecode: 0x142fa0f70>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

Some \proglang{R} packages for statistical learning include facilities
for cross-validation. Notable examples are the \pkg{caret} package
\citep{Kuhn:2008} and the \pkg{tidyfit} package \citep{Pfitzinger:2024};
the latter employs \pkg{rsample} \citep{FrickEtAl:2024} for CV, which we
will discuss presently. This contrasts with the approach taken in the
\pkg{cv} package, which, as we have explained, directly supports classes
of standard \proglang{R} statistical models.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library(rsample)
R> 
R> data("attrition", package = "modeldata")
R> nrow(attrition)
\end{CodeInput}
\begin{CodeOutput}
[1] 1470
\end{CodeOutput}
\begin{CodeInput}
R> # formula for model to be fit; Attrition is binary:
R> mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome
R> 
R> rs_obj <- loo_cv(attrition) # original uses vfold_cv()
R> 
R> # user-supplied function:
R> holdout_results <- function(splits, ...) {
+   mod <- glm(..., data=analysis(splits), family=binomial)
+   holdout <- assessment(splits)
+   res <- broom::augment(mod, newdata=holdout)
+   lvls <- levels(holdout$Attrition)
+   predictions <- factor(ifelse(res$.fitted > 0, lvls[2], lvls[1]),
+                         levels=lvls)
+   res$correct <- predictions == holdout$Attrition
+   res
+ }
R> 
R> library(purrr, warn.conflicts=FALSE)
R> # apply CV:
R> rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
R> # summarize results:
R> rs_obj$accuracy <- map_dbl(rs_obj$results,
+                            function(x) mean(x$correct))
R> # accuracy is the complement of the Bayes-rule CV criterion:
R> 1 - summary(rs_obj$accuracy) 
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 1.0000  0.0000  0.0000  0.1612  0.0000  0.0000 
\end{CodeOutput}
\end{CodeChunk}
